Synthesizing Verified Components for Cyber Assured Systems Engineering \thanks{DISTRIBUTION STATEMENT A. Approved for public release.}

### 2022-03-25 paper/agree-semantics.tex
Minor edits in section 4

We now give an overview of a formal model for AGREE. This provides a
setting in which we are able to relate the contract correctness
results discussed above, obtained via model-checking, with the code
generated from high-assurance contracts.  The syntax of AGREE is
essentially that of quantifier-free first order predicate logic
supplemented with a few temporal operators. The terms (\emph{e}) are
arithmetic expressions built from variables ($v$) and numeric and
boolean literals ($c$), while formulas (\emph{b}) are built using
logical connectives from atomic formulas (\emph{a}) based on the
familiar comparison operators.
\[
\begin{array}{rcl}
e & ::= & v \mid c \mid e \;\set{+,*,/}\; e \\
a & ::= & e\; \set{=,<}\; e \\
b & ::= & v \mid c \mid a \mid \neg b
            \mid b \; \set{\land,\lor,\imp,\iff}\; b
\end{array}
\]

There is also a conditional, $\itelse{b}{(-)}{(-)}$, for both terms
and formulas. Lastly, there are temporal operators $\konst{pre}(-)$,
\emph{delay} $(-) \to (-)$, and $\konst{Hist}(-)$.

The semantics of terms and formulas is in terms of \emph{streams of
values}. Values encompass at least booleans and numbers, but can be
readily extended to include records and arrays. A value stream is a
total function from time (natural numbers) to values:
\[
 \konst{stream} = \mathbb{N} \to \konst{value}
\]
Given an \emph{environment} $E : \konst{name} \mapsto \konst{stream}$
binding variable names to value streams, the semantics $\sem{-}^E_t$
of terms and formulas defines the meaning of compound syntax in terms
of the meaning of subexpressions. The value of a variable $v$ at time
$t$ is found by looking up the stream bound to $v$ in $E$ (call it
$s$) and returning $s_t$. Some clauses of the semantics follow,
omitting the temporal operators:
\[
\begin{array}{rcl}
\sem{v}^E_t & = & E(v)(t) \\
\sem{c}^E_t & = & c \\
\sem{e_1 + e_2}^E_t & = & \sem{e_1}^E_t + \sem{e_2}^E_t \\
   & \cdots & \\
\sem{b_1 \land b_2}^E_t & = & \sem{b_1}^E_t \land \sem{b_2}^E_t \\
   & \cdots & \\
\end{array}
\]

The temporal operators deal with time in more significant ways. The
value of $\konst{pre}(e)$ at time $t$ is the value of $e$ at time
$t-1$ (at time zero, \konst{pre} is undefined).  A delay $e_1 \to e_2$
temporally ``shifts'' $e_2$ by means of prepending the first element
of $e_1$ to it.

\[
\begin{array}{rcl}
\sem{\konst{pre}(e)}^E_t & = & \sem{e}^E_{t-1}, \mathrm{when}\ t > 0 \\
\sem{e_1 \to e_2}^E_t & = & \itelse{t=0}{\sem{e_1}^E_0}{\sem{e_2}^E_t} \\
\sem{\konst{Hist}(b)}^E_t & = & \forall n \leq t.\; \sem{b}^E_n
\end{array}
\]

Although basic, these definitions can be used to define higher-level
operators from PTLTL, such as \konst{Once} and \konst{Since}.

>>>>

\begin{comment}
This section gives an abstract mathematical overview of AGREE's
compositional reasoning system, defining assume-guarantee contracts
and the verification conditions that AGREE checks in order to
modularly establish overall system correctness at the model level.
The syntax and semantics of AGREE are sketched in some detail.  The
discussion then becomes more concrete, showing---by example---how the
AGREE contract language is used to specify the cyber-hardened system
of \figref{fig:hardened}. The discussion proceeds from the
system-level contract to the filter and monitor components, which rely
on the notion of a \emph{code contract} to support code generation.
\end{comment}

We now give an overview of a formal model for AGREE syntax and semantics.
The discussion then becomes more concrete, showing by example, how the AGREE language is used to specify the properties of the SW system introduced in \secref{sec:example}.
Any implementation of the SW system must guarantee the specification output under the system assumptions.

The syntax of AGREE is
essentially that of quantifier-free first order predicate logic
supplemented with a few temporal operators. The terms (\emph{e}) are
arithmetic expressions built from variables ($v$) and numeric and
boolean literals ($c$), while formulas (\emph{b}) are built using
logical connectives from atomic formulas (\emph{a}) based on the
familiar comparison operators.
\[
\begin{array}{rcl}
e & ::= & v \mid c \mid e \;\set{+,*,/}\; e \\
a & ::= & e\; \set{=,<}\; e \\
b & ::= & v \mid c \mid a \mid \neg b
            \mid b \; \set{\land,\lor,\imp,\iff}\; b
\end{array}
\]
There is also a conditional, $\itelse{b}{(-)}{(-)}$, for both terms
and formulas. Lastly, there are temporal operators $\konst{pre}(-)$ (i.e., \emph{yesterday} in PLTL),
\emph{followed-by} $(-) \to (-)$, and $\konst{Hist}(-)$.

The semantics of terms and formulas, as mentioned in the previous section, is in terms of \emph{streams of
values}. Values encompass at least booleans and numbers, but can be
readily extended to include records and arrays. A value stream is a
total function from time (natural numbers including 0) to values:
\[
 \konst{stream} = \mathbb{N}_0 \to \konst{value}
\]

Given an \emph{environment} $E : \konst{name} \mapsto \konst{stream}$
binding variable names to value streams, the semantics $\sem{-}^E_t$
of terms and formulas defines the meaning of compound syntax in terms
of the meaning of subexpressions. The value of a variable $v$ at time
$t$ is found by looking up the stream bound to $v$ in $E$ (call it
$s$) and returning $s_t$. Some clauses of the semantics follow,
omitting the temporal operators:
\[
\begin{array}{rcl}
\sem{v}^E_t & = & E(v)(t) \\
\sem{c}^E_t & = & c \\
\sem{e_1 + e_2}^E_t & = & \sem{e_1}^E_t + \sem{e_2}^E_t \\
   & \cdots & \\
\sem{b_1 \land b_2}^E_t & = & \sem{b_1}^E_t \land \sem{b_2}^E_t \\
   & \cdots & \\
\end{array}
\]

The temporal operators deal with time in more significant ways. The
value of $\konst{pre}(e)$ at time $t$ is the value of $e$ at time
$t-1$ (at time zero, \konst{pre} is undefined).  
A followed-by, $e_1 \to e_2$, evaluates to $e_1$ at time 0; otherwise it is $e_2$.
Followed-by is often used to give a value to a stream in the first step so that subsequent steps are able to use $\konst{pre}(e)$.
\[
\begin{array}{rcl}
\sem{\konst{pre}(e)}^E_t & = & \sem{e}^E_{t-1}, \mathrm{when}\ t > 0 \\
\sem{e_1 \to e_2}^E_t & = & \itelse{t=0}{\sem{e_1}^E_0}{\sem{e_2}^E_t} \\
\sem{\konst{Hist}(b)}^E_t & = & \forall n \leq t.\; \sem{b}^E_n
\end{array}
\]

Although basic, these definitions can be used to define higher-level
operators from PTLTL, such as \konst{Once} and \konst{Since}. 
\[
   \begin{array}{rcl}
      \sem{\konst{Once}(b)}^E_t & = & \sem{b \vee (\mathit{false} \to \konst{pre}(\konst{Once}(b)))}^E_t
   \end{array}
\]

### 2022-03-25 paper/paper.tex
re-re-org
Unify the temporal operators and clarify PLTL
Minor edits in section 4

\newcommand{\figref}[1]{Fig.~\ref{#1}}
\newcommand{\secref}[1]{Sec.~\ref{#1}}
\newcommand{\lineref}[1]{Line~\ref{#1}}
\newcommand{\linesref}[2]{Lines~\ref{#1}--\ref{#2}}

@@

@@

\subsection{SOSYM Journal Details}
The due date is \textbf{March} $22^\mathrm{nd}$, \textbf{2022}.
Go to \url{https://mc.manuscriptcentral.com/sosym/}. Create a new account (if you don't have one), login and go to the "Author"-section, select the link "Start New Submission" under Author Dashboard and select "Special Section Paper" as the paper type, and choose "MODELS 2021 Special Issue". Also, make sure to mention the MODELS 2021 special issue in the cover letter.

\subsection{Revisions}
\begin{compactitem}
  \item Revise the end of the abstract regarding the case studies to be consistent with what ends up in this version of the paper.
  \item Revise the end of the introduction that discusses the case studies to be consistent with what ends up in this version of the paper.
  \item Fix the link to the motivating example to point to the new repository once it is created. Fix in the next section too.
  \item Add discussion on the limitation of the approach: what attacks can and cannot be covered, can valid messages be rejected, etc.
\end{compactitem}

@@

\subsection{Revisions}
\begin{compactitem}
  \item \sout{Reformat text to put single sentence on each line.}
  \item Move example to its own repository, archive, and give the DOI and URL for the example.
  \item \sout{Update AGREE certificates with new model}
  \item Separate example into separate files for initial design, cyber-components and tests, hardened design, and seL4 process design.
  \item \sout{Make clear the initial properties for the assumed correct model:}
    \begin{compactitem}
      \item \sout{Assume at the system level that a new automation request is not sent unless the prior request, if any, is acknowledged}
      \item \sout{Assume at all levels input messages are well-formed}
      \item \sout{Guarantee at all levels that waypoint outputs coincide with air vehicle location messages}
      \item \sout{Guarantee that a start from the waypoint manager coincides with an automation response}
      \item \sout{Guarantee for all trusted components that outputs are well-formed}
    \end{compactitem}
  \item Add citations, and remove KLS footnote to last paragraph---should cite HAMR, seL4, and Adventium.
\end{compactitem}

\section{Compositional reasoning}
\label{sec:agree}
\input{agree}

\subsection{Revisions}
\begin{compactitem}
  \item \sout{Add in the AGREE specification for the system}
  \item \sout{Add in the AGREE verification conditions (e.g., similar to those in the Scheduled Components NFM submission)}
  \item Add reference to the Liskov principle of safe-substitution when stating that the system contract is a sound abstraction of the implementation.
  \item Add back in a few sentences on the contract for the filter since those have been removed with the addition of the formal definitions.
\end{compactitem}

\section{Formal contracts for the example system}
\input{sw-contract}

\subsection{Code contracts for cyber components}
\input{high-assurance-contracts}

@@

\subsection{Revisions}
\begin{compactitem}
  \item Add in intuitive definition or AGREE leaf-component semantics
  \item State well-formed theorem
  \item State correctness theorem or any key theorems to the synthesis proof
\end{compactitem}

@@

\subsection{Revisions}
\begin{compactitem}
  \item State theorem relating the step function to the meaning of the leaf-node semantics
\end{compactitem}

@@

\subsection{Revisions}
\begin{compactitem}
  \item Possible Outline:
  \begin{compactitem}
    \item Two case studies: phase 2 is the proof on concept on a industrial size problem while phase 3 tests if AGREE is expressive enough for stating complex properties and if designers are able to use it.
    \item Phase 2 write up focusses on the number of components and the relative ease for AGREE to prove out and for SPLAT to synthesize. It notes that the components were relatively simple.
    \item Phase 3 write up focusses on the shear complexity of the components and in particular the gate and the transport monitor. Here it the need for testing is is most obvious. Also obvious is the need to make AGREE more expressive and to separate AGREE into a specification and implementation language as seen in Dafny.
    \item Make the general observation that the transforms on the AADL model to add the components and connections doesn't really save much time when compared against the time it takes to write and test the specifications. The automations is nice, but not where time is saved. Where time is saved is in the assurance case because once a specification is done, it is synthesized and proved correct for ALL inputs and ALL outputs. Not just some. Not code inspections. Not additional test requirements. All that is saved.
  \end{compactitem}
  \item Clarify research questions answered by the case study
  \item Make the figures bigger
  \item Rewrite to focus on synthesized components and the AGREE specifications
  \item Add an interesting monitor from the TA6 model
  \item Discuss what efforts are optimized with the automations versus a fully manual process for creating the components--how does the automated transformation affect actual time for engineers
\end{compactitem}

@@

\subsection{Revisions}
\begin{compactitem}
  \item Add recent work published after 2018
\end{compactitem}

>>>>

\newcommand{\figref}[1]{Fig.~\ref{#1}}
\newcommand{\secref}[1]{Sec.~\ref{#1}}
\newcommand{\lineref}[1]{Line~\ref{#1}}
\newcommand{\linesref}[2]{Lines~\ref{#1}--\ref{#2}}
% \newcommand{\eqref}[1]{Eq.~\ref{#1}}

@@

\newif\ifREVISIONS
\REVISIONSfalse

@@

\ifREVISIONS
\subsection{SOSYM Journal Details}
The due date is \textbf{March} $22^\mathrm{nd}$, \textbf{2022}.
Go to \url{https://mc.manuscriptcentral.com/sosym/}. Create a new account (if you don't have one), login and go to the "Author"-section, select the link "Start New Submission" under Author Dashboard and select "Special Section Paper" as the paper type, and choose "MODELS 2021 Special Issue". Also, make sure to mention the MODELS 2021 special issue in the cover letter.
\fi

\ifREVISIONS
\subsection{Revisions}
\begin{compactitem}
  \item Revise the end of the abstract regarding the case studies to be consistent with what ends up in this version of the paper.
  \item Revise the end of the introduction that discusses the case studies to be consistent with what ends up in this version of the paper.
  \item Fix the link to the motivating example to point to the new repository once it is created. Fix in the next section too.
  \item Add discussion on the limitation of the approach: what attacks can and cannot be covered, can valid messages be rejected, etc.
\end{compactitem}
\fi

@@

\section{Compositional reasoning}
\label{sec:agree}
\input{verification-conditions}

\section{Syntax and semantics of AGREE}
\label{agree-semantics}
\input{agree-semantics}

\subsection{Formal contract for the example system}
\input{sw-contract}

\ifREVISIONS
\subsection{Revisions}
\begin{compactitem}
  \item \sout{Add in the AGREE specification for the system}
  \item \sout{Add in the AGREE verification conditions (e.g., similar to those in the Scheduled Components NFM submission)}
  \item \sotu{Add reference to the Liskov principle of safe-substitution when stating that the system contract is a sound abstraction of the implementation.}
  \item Add back in a few sentences on the contract for the filter since those have been removed with the addition of the formal definitions.
\end{compactitem}
\fi

\section{Code contracts for cyber components}
\label{code-contracts}
\input{high-assurance-contracts}

@@

\ifREVISIONS
\subsection{Revisions}
\begin{compactitem}
  \item Add in intuitive definition or AGREE leaf-component semantics
  \item State well-formed theorem
  \item State correctness theorem or any key theorems to the synthesis proof
\end{compactitem}
\fi

@@

\ifREVISIONS
\subsection{Revisions}
\begin{compactitem}
  \item State theorem relating the step function to the meaning of the leaf-node semantics
\end{compactitem}
\fi

@@

\ifREVISIONS
\subsection{Revisions}
\begin{compactitem}
  \item Possible Outline:
  \begin{compactitem}
    \item Two case studies: phase 2 is the proof on concept on a industrial size problem while phase 3 tests if AGREE is expressive enough for stating complex properties and if designers are able to use it.
    \item Phase 2 write up focusses on the number of components and the relative ease for AGREE to prove out and for SPLAT to synthesize. It notes that the components were relatively simple.
    \item Phase 3 write up focusses on the shear complexity of the components and in particular the gate and the transport monitor. Here it the need for testing is is most obvious. Also obvious is the need to make AGREE more expressive and to separate AGREE into a specification and implementation language as seen in Dafny.
    \item Make the general observation that the transforms on the AADL model to add the components and connections doesn't really save much time when compared against the time it takes to write and test the specifications. The automations is nice, but not where time is saved. Where time is saved is in the assurance case because once a specification is done, it is synthesized and proved correct for ALL inputs and ALL outputs. Not just some. Not code inspections. Not additional test requirements. All that is saved.
  \end{compactitem}
  \item Clarify research questions answered by the case study
  \item Make the figures bigger
  \item Rewrite to focus on synthesized components and the AGREE specifications
  \item Add an interesting monitor from the TA6 model
  \item Discuss what efforts are optimized with the automations versus a fully manual process for creating the components--how does the automated transformation affect actual time for engineers
\end{compactitem}
\fi

@@

\ifREVISIONS
\subsection{Revisions}
\begin{compactitem}
  \item Add recent work published after 2018
\end{compactitem}
\fi
